Forecasting the power of Higher Order Weak Lensing Statistics with automatically differentiable simulations

### 2022-03-01 paper/paper.tex

To understand how accurate our simulations are, and to validate the choice of the smoothing size to apply, we tried to reproduce the summary statistics from the only Dark $\kappa$TNG simulations. 
 We show the results from the angular power spectrum comparison for 3 source redshifts in Figure \ref{fig:clsktng_comp1arc}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/clsktng_comp1arc.png}
    \caption{
    \textbf{Upper panel}: Angular power spectra for 3 different source redshifts from
our DLL maps compared to the $\kappa$TNG predictions.
 \textbf{Lower panel}
The fractional differences of the DLL power spectra with respect to the $\kappa$TNG predictions.The shaded regions represent the standard deviation from 1000 $\kappa$TNG realisations. \\
 The power spectra mean over 100 map realisations.The maps are smoothed with a 1 arcmin Gaussian filter. }
    \label{fig:clsktng_comp1arc}
\end{figure}
% with a smoothing size $\theta=3.6$. We perform the angular power spectrum analysis in a range of $\ell=[300,3000]$ according to the requirement specified in  \cite{mandelbaum2018lsst}. 
\denise{draw conclusions from this plot}
For each map, we compute the power spectrum by using our own code implemented in the Tensorflow framework.
%and we take the average over 2000 maps realisation.

\subsubsection{Wavelet Peak Counts}
It has been shown that it's necessary to go beyond second-order statistics to fully capture the non-Gaussian information encoded in the peak of the matter distribution. Several studies have proved that the  weak-lensing peak counts provide a way to capture information from non-linear structures that it is complementary to the information extracted by power spectrum.
The peak counts identify region of weak lensing map where the density value is higher, in this way they rely to the massive structure.
There are different ways to record the weak lensing peaks: the number of peaks as a function of either the signal-to-noise ratio or the convergence $\kappa$ of the peaks. 
In this paper, we choose to bin the peaks as function of $\kappa$. 
As for the angular power spectrum, the weak lensing peaks are detected from the noise and smoothed simulated maps by using our own code.
%and by averaging over 2000 different realizations. 
A peak is identified as the pixel of larger value than its eight neighbors in the image, convolved with a starlet filter $\mathcal{W}(\theta_{\text{ker}})$:
\begin{equation}
    \textbf{Peaks}=(\kappa*\mathcal{W})(\theta_{\text{ker}})
\end{equation}
Similar to the power spectrum, we compare the peak counts statistic from our map to the one from the $\kappa$TNG.
This allow us to identify the bins to discard (bins where the distribution of peaks is not Gaussian) as well as, the wavelet scale to discard. 
In Figure 6, we show the number counts of peaks in the upper panels, and the fractional differences between the DLL and $\kappa$TNG-Dark maps in the lower panels.
Again we smooth the convergence maps with a 1 arcmin Gaussian filter. The peak are counted for 8 linearly spaced bins within the range $(\kappa*\mathcal{W})=[-0.2, 0.5]$. 
%The bins are chosen such that at least 30 peaks are recorded in each bin.

\ref{fig:peakktng_comp1arc}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/peakktng_comp1arc.png}
    \caption{
      \textbf{Upper panel}: Peak counts from our DLL map at source redshift $z_s=1.034$ compared to the $\kappa$TNG at the same redshift. \textbf{Lower panel}: The fractional differences between the DLL and $\kappa$TNG map.
 The peak counts mean over 100 map realisations. The maps are smoothed with a 1 arcmin Gaussian filter. }
    \label{fig:peakktng_comp1arc}
\end{figure}

\EiffL{Explain wavelets}

@@

\subsubsection{Wavelet $\ell_1-$norm}

\ref{fig:l1normktng_comp1arc}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/l1normktng_comp1arc.png}
    \caption{
      \textbf{Upper panel}: $l_1$norm from our DLL map at source redshift $z_s=1.034$ compared to the $\kappa$TNG at the same redshift. \textbf{Lower panel}: The fractional differences between the DLL and $\kappa$TNG map.
 The $l_1$norm mean over 100 map realisations. The maps are smoothed with a 1 arcmin Gaussian filter. }
    \label{fig:l1normktng_comp1arc}
\end{figure}

\EiffL{Explain what's the deal with this l1 norm}

@@

\subsection{Photometric redshift uncertainty}
Cosmic shear measurement require a large number of images, for such quantities, determining the redshift of the galaxies spectroscopically is not possible, for this reason, photometric redshift (''photo-z'') measurement are computed.
In this paper we use the Fisher formalism to investigate its impact on cosmological parameter estimation.
For our analysis we considered a Smail-type galaxy distribution p(z) (Smail et al. 1994),
\begin{equation}\label{photz}
    p(z) \propto z^{\alpha} e^{-(\frac{z}{z_0})^{\beta}}
\end{equation}

 We adopt  $\alpha=$, $\beta =$ and fix $z_0$ such that ?'

@@

The first step to introduce the IA term in our simulation, consists in computing the projected tidal field from the three-dimensional mass over-density field $\delta(\vec{x}$
as (\cite{harnois2021cosmic})
\begin{equation}
   \tilde{s}_{ij,2D}(\textbf{k}_{\bot})=
   2 \pi 	\left [ \frac{k_ik_j}{k^2}- \frac{1}{3} \right ]
   \tilde{\delta}_{2D}(\textbf{k}_{\bot})
   \mathcal{G}_{2D}(\sigma_g)
\end{equation}
where $\sigma_g$ is the smoothing scale of a two-dimensional smoothing kernel $\mathcal{G}_{2D}$, the tilde symbols $~$ refers to the Fourier transformed quantities, $\textbf{k}_{\bot}$ denotes the two Fourier wave-vector components perpendicular to the line of sight, and the indices $(i,j)$ represent the $x$ or $y$ component.
In this work we use a Gaussian kernel and set the values of $\sigma_g=(idk yet) h^{-1}$Mpc.
First of all, we computed the Fourier transform of the projected density field from the full boxes of the simulation. 
Then the smoothed tidal field is interpolated on the light-cone. 
Then, we couple the tidal field $s_{ij,2D}$ with the complex intrinsic ellipticities as (\cite{harnois2021cosmic}):
\begin{equation}
    \epsilon_{1}^{IA}=\frac{A_{IA\bar{C}_1\bar{\rho}(z)}}{D{z}} (s_{xx}-s_{yy}), \ \ 
      \epsilon_{2}^{IA}=-\frac{2A_{IA\bar{C}_1\bar{\rho}(z)}}{D{z}} s_{xy},
\end{equation}
(\denise{check this formula})
and compute the observed ellipticities as:
\begin{equation}
    \boldsymbol{\epsilon}^{obs}=
    \frac{\boldsymbol{\epsilon}^{int}+\textbf{g}}{1+\boldsymbol{\epsilon}^{int,*}\textbf{g}}, \ \  \text{with}  \ \
    \boldsymbol{\epsilon}^{int}=
    \frac{\boldsymbol{\epsilon}^{
     IA}+\boldsymbol{\epsilon^{ran}}}
     {1+\boldsymbol{\epsilon}^{IA,*}\boldsymbol{\epsilon^{ran}}}.
\end{equation}

>>>>

To understand how accurate our simulations are, and to validate the choice of the smoothing size to apply, we tried to reproduce the summary statistics from the only Dark $\kappa$TNG simulations. 
 We show the results from the angular power spectrum comparison for 3 source redshifts in Figure \ref{fig:clsktng_comp1arc}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/clsktng_comp1arc.png}
    \caption{
    \textbf{Upper panel}: Angular power spectra for 4 different smoothing size from
our DLL maps compared to the $\kappa$TNG predictions.
 \textbf{Lower panel}
The fractional differences of the DLL power spectra with respect to the $\kappa$TNG predictions. \\
 The power spectra mean over 100 map realisations.The maps are computed for a single source redshift $z=1.034$. }
    \label{fig:clsktng_comp1arc}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/clsktng_compsmoothing.png}
    \caption{
    \textbf{Upper panel}: Angular power spectra for 3 different source redshifts from
our DLL maps compared to the $\kappa$TNG predictions.
 \textbf{Lower panel}
The fractional differences of the DLL power spectra with respect to the $\kappa$TNG predictions.The shaded regions represent the standard deviation from 1000 $\kappa$TNG realisations. \\
 The power spectra mean over 100 map realisations.The maps are smoothed with a 1 arcmin Gaussian filter. }
    \label{fig:clsktng_compsmoothing}
\end{figure}
% with a smoothing size $\theta=3.6$. We perform the angular power spectrum analysis in a range of $\ell=[300,3000]$ according to the requirement specified in  \cite{mandelbaum2018lsst}. 
\denise{draw conclusions from this plot}
For each map, we compute the power spectrum by using our own code implemented in the Tensorflow framework.
%and we take the average over 2000 maps realisation.

\subsubsection{The wavelet Transform }

The Wavelet transform has been widely used in astronomical images analysis due to its ability to decompose astronomical data into components at different scales. 
This multiscale approach is well-suited for the study of astronomical data since their complex hierarchical structure. 
In this work we filter the original convergence maps with the Starlet transform, or the Isotropic Undecimated Wavelet Transform, which reconstructs an image $c_0$ as the sum of all the wavelet scales and the coarse resolution image $c_J$:
\begin{equation}
    c_0(x,y)=c_J(x,y)+\sum_{j=1}^{J_{max}} w_j(x,y)
\end{equation}

where $j = J_{max} + 1$ and $w_j$ is an image of the same size of the original one, whose represents the details at a given resolution scales $j$, corresponding to a spatial size of $2^j$ pixels.

 In this paper we chose a $J_{max}=5$, this Starlet filter applied to our map with a pixel size of 0.29 arcmin corresponds to a decomposition of the original map in 5 maps of resolution [0.59, 1.18 , 2.34, 4.69, 7.03] arcmin and a coarse map.

\subsubsection{Wavelet Peak Counts}
It has been shown that it's necessary to go beyond second-order statistics to fully capture the non-Gaussian information encoded in the peak of the matter distribution. Several studies have proved that the  weak-lensing peak counts provide a way to capture information from non-linear structures that it is complementary to the information extracted by power spectrum.
The peak counts identify region of weak lensing map where the density value is higher, in this way they rely to the massive structure.
There are different ways to record the weak lensing peaks: the number of peaks as a function of either the signal-to-noise ratio or the convergence $\kappa$ of the peaks. 
In this paper, 
%we choose to bin the peaks as function of $\kappa$. 
%As for the angular power spectrum, the weak lensing peaks are detected from the noise and smoothed simulated maps by using our own code.
%and by averaging over 2000 different realizations. 
a peak is identified as the pixel of larger value than its eight neighbors in the image, convolved with a starlet filter $\mathcal{W}(\theta_{\text{ker}})$:
\begin{equation}
    \textbf{Peaks}=(\kappa*\mathcal{W})(\theta_{\text{ker}}).
\end{equation}
Similar to the power spectrum, we compare the peak counts statistic from our map to the one from the $\kappa$TNG.
This allow us to identify the bins to discard (bins where the distribution of peaks is not Gaussian) as well as, the wavelet scale to discard. 
In Figure 6, we show the number counts of peaks in the upper panels, and the fractional differences between the DLL and $\kappa$TNG-Dark maps in the lower panels.
Again we smooth the convergence maps with a 1 arcmin Gaussian filter. The peak are counted for 8 linearly spaced bins within the range $(\kappa*\mathcal{W})=[-0.2, 0.5]$. 
%The bins are chosen such that at least 30 peaks are recorded in each bin.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/peakktng_comp1arc.png}
    \caption{
      \textbf{Upper panel}: Peak counts from our DLL map at source redshift $z_s=1.034$ compared to the $\kappa$TNG at the same redshift. \textbf{Lower panel}: The fractional differences between the DLL and $\kappa$TNG map.
 The peak counts mean over 100 map realisations. The maps are smoothed with a 1 arcmin Gaussian filter. }
    \label{fig:peakktng_comp1arc}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/peakktng_compallarc.png}
    \caption{
      The fractional differences between the DLL and $\kappa$TNG peak counts at source redshift $z_s=1.034$.
 The peak counts mean over 100 map realisations. The diferent panels show the results when the maps are smoothed with a 1, 2 and 3.6 arcmin Gaussian filter. }
    \label{fig:peakktng_compallarc}
\end{figure}

@@

\subsubsection{Wavelet $\ell_1-$norm}
It was shown in \cite{ajani2021starlet} that the ‘starlet $l_1$-norm’ has  several advantages compared to the standard peak and void statistics, i.e. 1) the $l_1-$norm includes the information encoded in all pixels of the map 2) avoids defining  what a peak is and what a void is. 
The ‘starlet $l_1$-norm’ is defined as the sum of the absolute values of the starlet coefficients of the weak lensing map in a given bin \textit{i}:
\begin{equation}
    l_1^{j,i}=\sum_{u=1}^{\text{coef}(\mathcal{S}_{j,i})} 
    |\mathcal{S}_{j,i}[u]|=
    ||\mathcal{S}_{j,i}||_1
\end{equation}
where the index \textit{u} runs to the total number of pixels in the \textit{i}-th bin and $\mathcal{S}_{j,i}$ is the set of all the wavelet coefficient computed at a given scale \textit{j}, such that $\mathcal{S}_{j,i}=
\left \{  w_{j,k}/B_i<w_{j,k}<B_{i+1}\right \}$, with \textit{k} the pixel index and $B_i$ and $B_{i+1}$ the bin edges.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/l1normktng_comp1arc.png}
    \caption{
      \textbf{Upper panel}: $l_1$norm from our DLL map at source redshift $z_s=1.034$ compared to the $\kappa$TNG at the same redshift. \textbf{Lower panel}: The fractional differences between the DLL and $\kappa$TNG map.
 The $l_1$norm mean over 100 map realisations. The maps are smoothed with a 1 arcmin Gaussian filter. }
    \label{fig:l1normktng_comp1arc}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/l1normktng_compallarc.png}
    \caption{
      The fractional differences between the DLL and $\kappa$TNG $l_{norm}$ at source redshift $z_s=1.034$.
 The $l_{norm}$ results mean over 100 map realisations. The different panels show the results when the maps are smoothed with a 1, 2 and 3.6 arcmin Gaussian filter. }
    \label{fig:l1normktng_compallarc}
\end{figure}

\subsection{Galaxies redshift distribution}
Cosmic shear measurement require a large number of images, for such quantities, determining the redshift of the galaxies spectroscopically is not possible, for this reason, photometric redshift (``photo-z'') measurement are computed.
In this paper we use the Fisher formalism to investigate its impact on cosmological parameter estimation.
In our analysis, according to \cite{zhang2021transitioning}, the underlying redshift distribution is modelled via the parametrized model
\begin{equation}\label{photz}
    \frac{dN}{dz} \propto z^{2} \exp{[-(z/z_0)^{\alpha}]},
\end{equation}

where we adopt $\alpha=0.94$ and fix $z_0=0.26$.

The redshift distributions for the fiducial source samples are shown in Figure . 
\begin{figure}\label{photoz}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/l1normktng_compallarc.png}
    \caption{
   Source sample redshift distributions for each tomographic bin. The y-axis shows the normalized \textit{n(z)} distributions such that $\sum n(z)dz=1.0$}
    \label{fig:l1normktng_compallarc}
\end{figure}

@@

\subsection{Photometric redshift uncertainty}

@@

The first step to introduce the IA term in our simulation, consists in computing the projected tidal field from the three-dimensional mass over-density field $\delta(\vec{x})$
as (\cite{harnois2021cosmic})
\begin{equation}
   \tilde{s}_{ij,2D}(\textbf{k}_{\bot})=
   2 \pi 	\left [ \frac{k_ik_j}{k^2}- \frac{1}{3} \delta_{ij}^k  \right ]
  \tilde{\delta}_{2D}(\textbf{k}_{\bot})
   \mathcal{G}_{2D}(\sigma_g)
\end{equation}
where $\sigma_g$ is the smoothing scale of a two-dimensional smoothing kernel $\mathcal{G}_{2D}$, the tilde symbols $\sim $ refers to the Fourier transformed quantities, $\textbf{k}_{\bot}$ denotes the two Fourier wave-vector components perpendicular to the line of sight, and the indices $(i,j)$ represent the $x$ or $y$ component.
In this work we use a Gaussian kernel and set the values of $\sigma_g=(idk yet) h^{-1}$Mpc.
First of all, we computed the Fourier transform of the projected density field from the full boxes of the simulation. 
Then the smoothed tidal field is interpolated on the light-cone. 
Then, the complex intrinsic ellipticities can be computed from the the tidal field $s_{ij,2D}$ as (\cite{harnois2021cosmic}):
\begin{equation}\label{complex_el}
    \epsilon_{1}^{IA}=- \frac{A_{IA}\bar{C}_1\bar{\rho}(z=0)}{D(z)} (s_{xx}-s_{yy}), \ \ 
      \epsilon_{2}^{IA}=-\frac{2A_{IA}\bar{C}_1\bar{\rho}(z=0)}{D(z)} s_{xy},
\end{equation}
and the observed ellipticities as:
\begin{equation}
    \boldsymbol{\epsilon}^{obs}=
    \frac{\boldsymbol{\epsilon}^{int}+\textbf{g}}{1+\boldsymbol{\epsilon}^{int,*}\textbf{g}}, \ \  \text{with}  \ \
    \boldsymbol{\epsilon}^{int}=
    \frac{\boldsymbol{\epsilon}^{
     IA}+\boldsymbol{\epsilon^{ran}}}
     {1+\boldsymbol{\epsilon}^{IA,*}\boldsymbol{\epsilon^{ran}}}.
\end{equation}
However, it is noted that the final expression of the intrinsic ellipticities will be:
\begin{align}
    \tilde{ \epsilon}_{1}^{IA}(\textbf{k}_{\bot}) & \propto
    \left ( \frac{k_x^2-k_y^2}{k^2} \right )  \tilde{\delta}_{2D}(\textbf{k}_{\bot})
   \mathcal{G}_{2D}(\sigma_g) \\
       \tilde{ \epsilon}_{1}^{IA}(\textbf{k}_{\bot}) & \propto
    \left ( \frac{k_x k_y}{k^2} \right )  \tilde{\delta}_{2D}(\textbf{k}_{\bot})
   \mathcal{G}_{2D}(\sigma_g) \nonumber
\end{align}
that, except for the smoothing kernel, 
are the same filters used to produce shear maps from the convergence maps by applying the Kaiser  Squires (1993, KS hereafter) inversion.
Therefore, instead of using the Eq.\ref{complex_el} and applying the Kaiser Squires inversion to derive the $\kappa_{IA}$, 
we decided to implement the model on the convergence map level, i.e. calculate the intrinsic alignment as part of the convergence map: 
\begin{equation}
    \kappa_{IA_{i}}= 
    - A_{IA}\bar{C}_1\bar{\rho}(z=0)
    \int_{z_{min}}^{z_{max}} 
    n_{i}(z) \kappa_{z_{i}}
    \frac{ dz}{D(z)}
\end{equation}
where the index \textit{i} refers to the \textit{i}-th redshift bins.
\denise{not sure the part inside the integral, because D(z) was calculated before the integral}
